Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-12-11 scagnostics.Rmd
Replacing Fig. by Figure in all text.

Given the results from Fig.\ref{fig:plannedtour} we expect to be able to efficiently identify the sine view in the 6-d distribution, using only method 3 for the optimization, using e.g. the splines2d index. Indeed the guided tour quickly converges in this case, the evolution of the index functions is shown in Fig \ref{fig:findsine}. The only difficulties stem from the rotation dependence of the index function as well as the structured distribution, this will be discussed in some more details below. Here we only note that even for anchor planes highlighted by blue vertical lines the index sometimes decreases because of this dependence. These are however not dramatic effects because this search method is closely related to how interpolation is done (a different picture would be obtained using method 2 which often leads to large drops in index values after the interpolation step).

@@

The spiral is the most challenging structure to detect amongst the distribution considered here (compare also Posses results?). Here we explore the role of dimensionality in the index optimization, using the skinny index that was found to be sensitive to this structure.
We use a setup similar to that introduced for detecting the pipe structure, i.e.\ we use the search\_better optimization of the guided tour, disabling the cooling and with a large search window, while sampling a large number of random projections (up to 5000) for each step.
We consider distributions in d=4,6,8 dimensions and compare features found in the tour.
In d=4 dimensions the optimization identifies a view showing the spiral after 33 steps, and we limit the optimizer to consider only 33 steps for the study in 6 and 8 dimensions for direct comparison. Note however that the spiral is typically also not identified with much larger number of steps, the limiting factor being rather the number of random projections sampled and rotation non-invariance of the index prohibiting the convergence, in particular in higher dimensions.
Results are shown in the top panel Fig.~\ref{fig:findspiral}, showing the pairwise distance of the 33 planes found in 4, 6 and 8 dimensions (left), and comparing the evolution of the distance from the plane containing the spiral view with progression of the optimization (right). We see that the pairwise distances grow with d (what do we learn from that?). Moreover we find that the distance to the special view is far from monotonically decreasing towards zero, and large jumps are possible. Note that the spiral itself is approximately rotation invariant, but partial views are not. At medium distances from the special view we often observe a jump to higher distances as consequence of in-plane rotation dependence for such partial views. This stabilizes closer to the special view, however it relies on randomly identifying a projection within this angle. As a consequence increasing dimensionality dramatically reduces the chance of identifying structures with high index values.

@@

Fig. \ref{fig:bbhGuided3} shows six maxima obtained by different starts, for two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Four of the six (a-d) are almost identical, but not interesting projections. Projection f has the highest PP index value but it is primarily the view seen in the bivariate plot of dec and ra.

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Fig. \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{constructedExample}. Manually reconstructing this can be done by differencing the two parameters, in their original units. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precision angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}

>>>>

Given the results from Figure\ref{fig:plannedtour} we expect to be able to efficiently identify the sine view in the 6-d distribution, using only method 3 for the optimization, using e.g. the splines2d index. Indeed the guided tour quickly converges in this case, the evolution of the index functions is shown in Fig \ref{fig:findsine}. The only difficulties stem from the rotation dependence of the index function as well as the structured distribution, this will be discussed in some more details below. Here we only note that even for anchor planes highlighted by blue vertical lines the index sometimes decreases because of this dependence. These are however not dramatic effects because this search method is closely related to how interpolation is done (a different picture would be obtained using method 2 which often leads to large drops in index values after the interpolation step).

@@

The spiral is the most challenging structure to detect amongst the distribution considered here (compare also Posses results?). Here we explore the role of dimensionality in the index optimization, using the skinny index that was found to be sensitive to this structure.
We use a setup similar to that introduced for detecting the pipe structure, i.e.\ we use the search\_better optimization of the guided tour, disabling the cooling and with a large search window, while sampling a large number of random projections (up to 5000) for each step.
We consider distributions in d=4,6,8 dimensions and compare features found in the tour.
In d=4 dimensions the optimization identifies a view showing the spiral after 33 steps, and we limit the optimizer to consider only 33 steps for the study in 6 and 8 dimensions for direct comparison. Note however that the spiral is typically also not identified with much larger number of steps, the limiting factor being rather the number of random projections sampled and rotation non-invariance of the index prohibiting the convergence, in particular in higher dimensions.
Results are shown in the top panel Figure \ref{fig:findspiral}, showing the pairwise distance of the 33 planes found in 4, 6 and 8 dimensions (left), and comparing the evolution of the distance from the plane containing the spiral view with progression of the optimization (right). We see that the pairwise distances grow with d (what do we learn from that?). Moreover we find that the distance to the special view is far from monotonically decreasing towards zero, and large jumps are possible. Note that the spiral itself is approximately rotation invariant, but partial views are not. At medium distances from the special view we often observe a jump to higher distances as consequence of in-plane rotation dependence for such partial views. This stabilizes closer to the special view, however it relies on randomly identifying a projection within this angle. As a consequence increasing dimensionality dramatically reduces the chance of identifying structures with high index values.

@@

Figure \ref{fig:bbhGuided3} shows six maxima obtained by different starts, for two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Four of the six (a-d) are almost identical, but not interesting projections. Projection f has the highest PP index value but it is primarily the view seen in the bivariate plot of dec and ra.

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Figure \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{constructedExample}. Manually reconstructing this can be done by differencing the two parameters, in their original units. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precision angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}
